---
title: "Best Coding Practices -- tidy, neat and efficient code"
subtitle: "RaukR 2019 • Advanced R for Bioinformatics"
author: "Marcin Kierczak"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: united
    highlight: textmate
    df_print: paged
    code_folding: none
    self_contained: false
    keep_md: false
    encoding: "UTF-8"
    css: ["assets/lab.css"]
---

```{r,echo=FALSE,child="assets/header-lab.Rmd"}
```


```{r,echo=FALSE,results='hide'}
## LIBRARIES AND VARIABLES
# load the packages you need for this document

#library(ggplot2)
#library(tidyverse)
library(bsplus)

# variables needed in this document
# raukr colours
#rv_col_dark <- "#125687"
#rv_col_light <- "#e7eef3"
```

<img src="assets/logo.svg" alt="logo_raukr" class="trlogo">

---

<p class="abstract">
The objective of this lab is to improve your coding skills by focusing on coding style, code benchmarking and optimization. Below, you will find a number of tasks connected to the topics covered in the *Best Coding Practices* lecture. Some tasks extend lectures content and require you to find some more information online. Please, note that while we are providing example solutions to many tasks, these are only **examples**. If you solve a task in a different way it does not matter your solution is wrong. In fact, it may be better than our solution. If in doubt, ask PI for help. We are here for you!
</p>

---

# Coding Style

## Task: Valid Variable Names.
Which of the following are valid/good variable names in R. What is wrong with the ones that are invalid/bad?
`var1`, `3way_handshake`, `.password`, `__test__`, `my-matrix-M`, `three.dimensional.array`, `3D.distance`, `.2objects`, `wz3gei92`, `next`, `P`, `Q`, `R`, `S`, `T`, `X`, `is.larger?`

## Task: Obscure Code.  
The code below works, but can be improved. Do improve it!
```
myIterAtoR.max <- 5
second_iterator.max<-7
col.NUM= 10
row.cnt =10
fwzy45 <- matrix(rep(1, col.NUM*row.cnt),nrow=row.cnt)
for(haystack in (2-1):col.NUM){
  for(needle in 1:row.cnt) {
if(haystack>=myIterAtoR.max){
fwzy45[haystack, needle]<-NA}
}}
```

```{r, answer=T, eval=FALSE}
iter_max <- 5
col_num <- 10
row_num <- 10
A <- matrix(rep(1, col_num * row_num), nrow = row_num)
for (i in 1:col_num) {
  for (j in 1:row_num) {
    if (i >= iter_max) {
      A[i, j] <- NA
    }
  }
}

# Can you improve the code more by eliminating loops or at least one of them?
```

## Task: Better Formatting.
Improve formatting and style of the following code:
```
simulate_genotype <- function( q, N=100 ) {
  if( length(q)==1 ){
    p <- (1 - q)
    f_gt <- c(p^2, 2*p*q, q^2) # AA, AB, BB
  }else{
    f_gt<-q
  }
  tmp <- sample( c('AA','AB','BB'), size =N, prob=f_gt, replace=T )
  return(tmp)
}
```

```{r, answer=T, eval=FALSE}
simulate_genotype <- function(q, N = 100) {
  if (length(q) == 1) {
    p <- (1 - q)
    f_gt <- c(p^2, 2*p*q, q^2) # AA, AB, BB
  } else {
    f_gt <- q
  }
  genotype <- sample(c('AA', 'AB', 'BB'), 
                     size = N, 
                     prob = f_gt, 
                     replace = T)
  return(genotype)
}
```

## Task: Hidden Variable.
Assign a vector of three last months (abbreviated in English) in a year to a hidden variable `my_months`.
```{r, answer=T, eval=FALSE}
.my_months <- rev(rev(month.abb)[1:3])
```

## Task: Pipeline-friendly Function.
Modify the function below so that it works with R pipes `%>%`:
```
my_filter <- function(threshold = 1, data, scalar = 5) {
  data[data >= threshold] <- NA 
  data <- data * scalar
  return(data)
}
```
 
**Note** you need to have the `magrittr` or `tidyverse` package loaded in order to be able to use the pipe `%>%`!
 
```{r answer=T, eval=FALSE}
my_filter <- function(data, threshold = 1, scalar = 5) {
  data[data >= threshold] <- NA 
  data <- data * scalar
  return(data)
}

# Test:
c(-5, 5) %>% my_filter()
```

## Task: Untidy Code?
Is the code below correct? Can it be improved?
```
simulate_phenotype <- function(pop_params, gp_map, gtype) {
  pop_mean <- pop_params[1]
  pop_var <- pop_params[2]
  pheno <- rnorm(n = N, mean = pop_mean, sd = sqrt(pop_var))
  effect <- rep(0, times = length(N))
  for (gt_iter in c('AA', 'AB', 'BB')) {
    effect[gtype == gt_iter] <- rnorm(n = sum(gtype == gt_iter), 
                                      mean = gp_map[gt_iter, 'mean_eff'], 
                                      sd = sqrt(gp_map[gt_iter, 'var_eff']))
  }
  dat <- data.frame(gt = gtype, raw_pheno = pheno, effect = effect, pheno = pheno + effect)
  return(dat)
}
```

```{r eval=FALSE, answer=T}
Maybe some small improvements can be done, but in principle the code is clean! Except that... the N is not initialized anywhere.
```

# Structuring the Code
## Task: Computing Variance.
Write a modular code (function or functions) that computes the sample standard deviation given a vector of numbers. Decide how to logically structure the code. Assume there are no built-in R functions for computing mean and variance available. The formula for variance is: $SD = \sqrt{\frac{\Sigma_{i=1}^{N}(x_i - \bar{x})^2}{N-1}}$. Standard deviation is $Var=SD^2$.  
<br>
**Hint:** consider that you may want to re-use some computed values in future, e.g. variance. 

```{r answer=T, eval=FALSE}
sample_mean <- function(x) {
  mean <- sum(x) / length(x)
  return(mean)
} 

sum_squared_deviations <- function(x) {
  tmp <- (x - sample_mean(x)) ^ 2
  sum_sq_dev <- sum(tmp)
  return(sum_sq_dev)
}

std_dev <- function(x) {
  variance <- sqrt(sum_squared_deviations(x) / (length(x) - 1))
  return(variance)
}

variance <- function(x) {
  return(std_dev(x) ^ 2)
}
```

## Task: Writing a Wrapper Function.
You found two functions in two different packages: the `randomSampleInt` function that generates a random sample of integer numbers and the `randomSampleLetter` function for generating a random sample of letters. Unfortunately, the functions are called in different ways which you want to unify in order to use them interchangeably in your code. Write a wrapper function around the `randomSampleLetter` that will provide the same interface to the function as the `randomSampleInt`. Also, the `randomSampleLetter` cannot handle the seed. Can you add this feature to your wrapper?

```
randomSampleInt <- function(x, verbose, length, seed = 42) {
  if (verbose) {
    print(paste0('Generating random sample of ', length, ' integers using seed ', seed))
  }
  set.seed(seed)
  sampleInt <- sample(x = x, size = length, replace = TRUE)
  return(sampleInt)
} 

randomSampleLetter <- function(N, silent=T, lett) {
  if (!silent) {
    print(paste0('Generating random sample of ', N, ' letters.'))
  }
  sample <- sample(x = lett, size = N, replace = TRUE)
  return(sample)
}
```

```{r, answer=T, eval=FALSE}
randomSampleLetterWrapper <- function(x, verbose, length, seed = 42) {
  set.seed(seed)
  result <- randomSampleLetter(N = length, silent = !verbose, lett = x)
  return(result)
}
```

## Task: Customizing `plot`.
Write a wrapper around the `graphics::plot` function that modifies its default behaviour so that it plots red crosses instead of black points. Do it in a way that enables the user to modify other function arguments. **Hint:** you may want to have a look at `graphics::plot.default`.

```{r, answer=T, eval=F}
my_plot <- function(x, ...) {
  plot(x, pch = 3, color = 'red', ...) 
}
```

## Task: Adding Arguments to a Function.
What if you want to pass some additional parameters to a function and, sadly, the authors forgot to add `...` to the list of function arguments. There is a way out -- you can bind extra arguments supplied as `alist` structure to the original function arguments retrieved by `formals`. Try to fix the function below, so that the call `red_plot(1, 1, col='red', pch=19)` will result in points being represented by red circles. Do use `alist` and `formals` and do not edit the `red_plot` itself! **Hint:** read help for `alist` and `formals`. Original function:
```
red_plot <- function(x, y) { 
  plot(x, y, las=1, cex.axis=.8, ...)
}
```

```{r answer=T, eval=F}
red_plot <- function(x, y) { 
  plot(x, y, las=1, cex.axis=.8, ...)
}

red_plot(1, 1, col='red', pch=19) # Does not work.
formals(red_plot) <- c(formals(red_plot), alist(... = )) # Fix.
red_plot(1, 1, col='red', pch=19) # Works!
```

## Task: Using `options`.
Use `options` to change the default prompt in R to `hello :-) > `.
Check what options are stored in the **hidden** variable called *Options*. 

```{r answer=T, eval=F}
options(prompt = "hello :-) > ")
.Options
options(prompt = "> ") # restoring the default
```

# Debugging
## Task: Code Correctness
Which of the following chunks of code are correct and which contain errors? Identify these errors.

### Chunk 1
```
input <- sample(1:1000, size = 1000, replace = T)
currmin <- NULL
for (i in input) {
  if (input > currmin) {
    currmin <- input
    print(paste0("The new minimum is: ", currmin))
  }
}
```

### Chunk 2
```
input <- sample(1:1000, size = 1000, replac = T)
currmin <- NULL
for (i in input) {
  if (input < currmin) {
    currmin <- input
    print(paste0("The new minimum is: ", currmin))
  }
}
```

### Chunk 3
```
for (cnt in 1:100) {
  if (cnt > 12) {
    print("12+")
  } else {
    print("Not 12+")
  }
}
```

### Chunk 4
```
result <- logical(10)
input <- sample(1:10, size = 10, replace = T)
for (i in 0:length(input)) {
  if (input[i] >= 5) {
    result[i] <- TRUE
  }
}
```

## Task: Debugger.
Play with debugger as described in lecture slides 17-21. 

## Task: Floating-point Arithmetics.
Can you fix the code below so that it produces more reliable result?
<br>
**Hint:** think in terms of system-specific representation $\epsilon$.  
<br>
Put the value of your double $\epsilon$ into [this spreadsheet](https://docs.google.com/spreadsheets/d/1_2tDeEkDVS06RkB437yBI1XEB5SUebtHWyxAf_aRJu4/edit?usp=sharing) (Best Coding Practises Lab sheet).
```
vec <- seq(0.1, 0.9, by=0.1)
vec == 0.7
```

```{r answer=T, eval=FALSE}
# One way is to use epsilon
# Check machine's floating point representation
vec <- seq(0.1, 0.9, by=0.1)

# Make a custom function that uses machines' epsilon for comparing
# values
is_equal <- function(x, y) {
  isEqual <- F
  if (abs(x - y) < unlist(.Machine)['double.eps']) {
    isEqual <- T
  }
  isEqual
}

# Some tests
0.7 == 0.6 + 0.1
is_equal(0.7, 0.6 + 0.1)
0.7 == 0.8 - 0.1
is_equal(0.7, 0.8 - 0.1)

# Now you can use the is_equal to fix the code!
```

# Profiling

## Task: Filling A Large Matrix.
Create a 10 000 x 10 000 matrix and fill it with random numbers (from 1 to 42), first row by row and later column by column. Use `proc.time` to see if there is any difference. Is the measurement reliable? Record the values you got in this spreadsheet: 

```{r eval=FALSE, answer=T, cache=TRUE}
N <- 10e3 * 10e3

# By row
t1 <- proc.time()
M <- matrix(sample(1:42, size = N, replace = T), nrow = sqrt(N), byrow = T)
t2 <- proc.time()
(t2 - t1)

# By column
t1 <- proc.time()
M <- matrix(sample(1:42, size = N, replace = T), nrow = sqrt(N), byrow = F)
t2 <- proc.time()
(t2 - t1)
```

## Task: Timing Reliability.
In the lecture slides, you have seen how to time sampling from Gaussian distribution:
```
system.time(rnorm(n = 10e6))
```
Is such single measurement reliable? Run the code 100 times, plot and record the mean and the variance of the `elapsed` time. Put these values (elapsed.time mean and variance) into [this spreadsheet](https://docs.google.com/spreadsheets/d/1_2tDeEkDVS06RkB437yBI1XEB5SUebtHWyxAf_aRJu4/edit?usp=sharing) (Best Coding Practises Lab sheet).

```{r answer=T, cache=T, eval=FALSE}
timing <- double(100)
for (i in 1:100) {
  st <- system.time(rnorm(n = 10e6))
  timing[i] <- st[3]
}
boxplot(timing) 
mean(timing)
var(timing)
```
An alternative approach or, more exactly, an alternative notation that achieves the same as the previous chunk of code but in a more compact way makes use of the `replicate`, a wrapper function around `sapply` that simplifies repeated evaluation of expressions. The drawback is you do not get the vector of the actual timing values but the results of calling `system.time` are already averaged for you. Try to read about the `replicate` and use it to re-write the code above. Put the `elapsed.time` into [this spreadsheet](https://docs.google.com/spreadsheets/d/1_2tDeEkDVS06RkB437yBI1XEB5SUebtHWyxAf_aRJu4/edit?usp=sharing) (Best Coding Practises Lab sheet). How does this value compare to calling `system.time` within a loop in the previous chunk of code? Are the values similar?

```{r answer=T, cache=T, eval=FALSE}
st2 <- system.time(replicate(n = 100, rnorm(n = 10e6)))
```

## Task: Microbenchmarking.
While `system.time` might be sufficient most of the time, there is also an excellent package `microbenchmark` that enables more accurate time profiling, aiming at microsecond resolution that most of modern operating systems offer. Most of the benchmarking the `microbenchmark` does is implemented in low-overhead C functions and also the package makes sure to:
* estimate granularity and resolution of timing for your particular OS,
* warm up your processor before measuring, i.e. wake the processor up from any idle state or likewise. 

Begin by installing the `microbenchmark` package.

### Checking System Time.
Check the current value of the platform's timer.
```{r answer=T, eval=F}
microbenchmark::get_nanotime()
```
Modify the code below so that it uses the current value of platform's timer:
```
timing <- double(100)
for (i in 1:100) {
  st <- system.time(rnorm(n = 10e6))
  timing[i] <- st[3]
}
boxplot(timing)
```
Put the mean and the variance into [this spreadsheet](https://docs.google.com/spreadsheets/d/1_2tDeEkDVS06RkB437yBI1XEB5SUebtHWyxAf_aRJu4/edit?usp=sharing) (Best Coding Practises Lab sheet, Microbenchmark -- loop)
```{r answer=T, eval=F}
library(microbenchmark)
timing <- double(100)
for (i in 1:100) {
  nanotime_start <- get_nanotime()
  rnorm(n = 10e6)
  nanotime_stop <- get_nanotime()
  timing[i] <- nanotime_stop - nanotime_start
}
mean(timing)
var(timing)
boxplot(timing)
```

### Microtiming Precision.
There is an experimental function in the `microbenchmark` package that helps the package estimate granularity and resolution of your particular timing subsystem. According to the documentation, *the function measures the overhead of timing a C function call rounds times and returns all non-zero timings observed.*
<br>
Run the `microtiming_precision` function and put the mean and the variance of the resulting vector into [this spreadsheet](https://docs.google.com/spreadsheets/d/1_2tDeEkDVS06RkB437yBI1XEB5SUebtHWyxAf_aRJu4/edit?usp=sharing) (Best Coding Practises Lab sheet, Microbenchmark -- precision)
```{r answer=T, eval=F}
precision <- microbenchmark::microtiming_precision()
mean(precision)
var(precision)
```
Run the function one time without assigning its value to a variable and consult the documentation. Compare the output of running the function without assigning the value to a variable, the values stored in the variable by the function upon assignment and the value specified in the documentation.
```{r answer=T, eval=F}
# In version 1.4-4 of the package, all three ways give different results!
microbenchmark::microtiming_precision()
precision <- microbenchmark::microtiming_precision()
?microbenchmark::microtiming_precision
```

### The Microbenchmark Way.
Finally, let's benchmark our `rnorm` example using `microbenchmark`: 

* microbenchmark the `rnorm(n = 10e6)` expression,
* plot the results using both `ggplot2` and a boxplot (read the `microbenchmark` package documentation),
* look at the summary of the benchmark,
* how long does it take to dispatch a simple function that does nothing compared to evaluating a constant and adding two integers?

```{r answer=TRUE, eval=F}
require(microbenchmark)
require(ggplot2)
mb <- microbenchmark(rnorm(n = 10e6))
autoplot(mb)
boxplot(mb)
summary(mb)
f <- function() {}
mb2 <- microbenchmark(f(), pi, 2+2)
summary(mb2)
autoplot(mb2)
```

## Task: More Advanced Profiling.

Now, we will use a even more sophisticated approach to profiling. 

### The `Rprof` way.
* Write three functions that fill by row a $N \times N$ matrix $M$ with randomly generated numbers from a vector given as argument `bag`, allow for passing random seed value as function argument with the default value of 42. After filling the matrix with values, add to each and every element of $M$ the number of column the element is in and return such matrix from the function. Functions should: 

* `fill_alloc`) -- use memory allocation prior to loop in which the matrix is being filled and allocate memory using `init` value passed as argument and by default set to `NULL`,
* `fill_noalloc` -- not use memory allocation prior to the loop,
* `fill_noloop` should not the loop for filling the matrix in. 

** NOTE! ** do not perform addition of column number in the same 

Following this and using `rnorm(1000, mean = 0, sd = 1)`:

* use `Rprof` to profile the functinos using the same seed and N=100,
* use `Rprof` to check whether there is a difference between initializing the matrix using `NULL` and 0 in `fill_alloc`,
* what happens if $N = 10$ compared to $N = 20$ to $N = 100$? 

```{r answer=T, eval=F}
fill_noloop <- function(N, bag, seed = 42) {
  set.seed(seed)
  values <- sample(bag, size = N^2, replace = T)
  M <- matrix(data = values, nrow = N, byrow = T)
  for (col_num in 1:N) {
    M[, col_num] <- M[, col_num] + col_num
  }
  return(M)
}

fill_noalloc <- function(N, bag, seed = 42) {
  set.seed(seed)
  values <- sample(bag, size = N^2, replace = T)
  M <- NULL
  cnt = 1
  for (row in 1:N) {
    row_tmp <- c()
    for (col in 1:N) {
      row_tmp <- c(row_tmp, values[cnt])
      cnt <- cnt + 1
    }
    M <- rbind(M, row_tmp)
  }
  for (col_num in 1:N) {
    M[, col_num] <- M[, col_num] + col_num
  }
  return(M)
}

fill_alloc <- function(N, bag, seed = 42, init = NA) {
  set.seed(seed)
  values <- sample(bag, size = N^2, replace = T)
  M <- matrix(rep(init, times=N^2), nrow = N, byrow = T)
  cnt = 1
  for (row in 1:N) {
    for (col in 1:N) {
      M[row, col] <- values[cnt]
      cnt <- cnt + 1
    }
  }
  for (col_num in 1:N) {
    M[, col_num] <- M[, col_num] + col_num
  }
  return(M)
}

summary <- summaryRprof('profiler_test_fillers.out', memory='both')
summary$by.self

# answers to the remaining questions are not given
```

## Optimization
Have a look at our answers. 

* How can you optimize the `fill_alloc` even further (call the optimized version `fill_alloc_opt`)?

```{r answer = T, eval=F}
fill_alloc_opt <- function(N, bag, seed = 42, init = NA) {
  set.seed(seed)
  values <- sample(bag, size = N^2, replace = T)
  M <- matrix(rep(init, times=N^2), nrow = N, byrow = T)
  cnt = 1
  for (row in 1:N) {
    for (col in 1:N) {
      M[row, col] <- values[cnt] + col
      cnt <- cnt + 1
    }
  }
  return(M)
}
```

* Optimize the `fill_noloop` to `fill_noloops` that does not use any loops at all.

```{r answer=T, eval=F}
fill_noloops <- function(N, bag, seed = 42) {
  values <- sample(bag, size = N^2, replace = T)
  inc <- rep(x = 1:N, times = N)
  M <- matrix(data = values + inc, nrow = N, byrow = T)
  return(M)
}
```

## Using the `profr` package.
* Install and load the `profr` package.
* Use `profr` to profile `fill_noloop`, `fill_noloops` and `fill_alloc_opt`.

```{r answer=T, eval=F}
library(profr)
Rprof("profr_noloop.out", interval = 0.01)
fill_noloop(1000, rnorm(1000), seed = 42)
Rprof(NULL)
profile_noloop_df <- parse_rprof('profr_noloop.out')

Rprof("profr_noloops.out", interval = 0.01)
fill_noloops(100, rnorm(1000), seed = 42)
Rprof(NULL)
profile_noloops_df <- parse_rprof('profr_noloops.out')

Rprof("profr_alloc_opt.out", interval = 0.01)
fill_alloc_opt(10, rnorm(1000), seed = 42)
Rprof(NULL)
profile_alloc_opt_df <- parse_rprof('profr_alloc_opt.out')

profr::ggplot.profr(profile_noloop_df)
profr::ggplot.profr(profile_noloops_df)
profr::ggplot.profr(profile_alloc_opt_df)
```

## Using the `profvis` package.
* Install and load the `profvis` package.
* Use `profvis` to profile `fill_noloop`, and `fill_alloc` functions.


# Optimize Your Code

In this section, we will deal with some selected ways to optimize your code.

## Task: Optimize This!
Given is a function:
```
optimize_me <- function(N = 1000, values = c(1:1e4)) {
  N = 10; values = c(1:1e4)
  dat1 <- matrix(size = N^2)
  for (i in 1:N) {
    for (j in 1:N) {
      dat1[i, j] <- sample(values, 1)
    }
  }
  dat0 <- dat1
  dat1[lower.tri(dat1)] <- t(dat1)[lower.tri(dat1)]
  
  dat2 <- NULL 
  for (i in 1:N) {
    i_tmp <- c()
    for (j in 1:N) {
      i_tmp <- c(i_tmp, sample(values, 1))
    }
    dat2 <- rbind(dat2, i_tmp)
  }
  dat2[lower.tri(dat2)] <- t(dat2)[lower.tri(dat2)]
 
  M <- dat2
  for (i in 1:N) {
    for (j in 1:N) {
      M[i, j] <- dat1[i, j] * dat2[i, j]
    }
  }
  for (i in 1:N) {
    for (j in 1:N) {
      M[i, j] <- M[i, j] + values[3]
    }
  }
  N <- M %*% dat0
  result <- apply(N, 2, mean)
  return(result)
}
```

* What does it do, step-by-step?
* Profile it.
* Is `dat1 <- matrix(size = N^2)` better than `dat1 <- matrix(NA, nrow=N, ncol=N)`?
* Can you optimize something using `BLAS`?
* Can you optimize by using `apply` somewhere?
* Can you optimize `apply` further?
* What else can you optimize. Do it. Report speed gain and memory gain compared to the original version in [this spreadsheet](https://docs.google.com/spreadsheets/d/1_2tDeEkDVS06RkB437yBI1XEB5SUebtHWyxAf_aRJu4/edit?usp=sharing) (Best Coding Practises Lab sheet, Optimization gains).  

```{r, echo=FALSE, child="assets/session.Rmd"}
```

```{r, echo=FALSE, child="assets/build.Rmd"}
```

```{r, echo=FALSE, child="assets/footer.Rmd"}
```
